---
title: "Chpt. 14: Strings & Regex (regular expressions)"
output: html_notebook
---



```{r}
x <- "this is a string"
x
class(x)
y <- "3"
class(y)
as.numeric(y)
as.numeric(x) #doesn't work! 

z <- c("I", "am", "a", "string", "vector")
z
z[1:2]
z[2:5]
z[-5]

#ugly
print(z)
#simplified and interpreted (pretty)
cat(z)

```

escape special characters with \
```{r}
a <- "As Michael Scott said, \"I'm not superstitious, but I am a little stitious.\""
print(a)
cat(a)
```
line break with \n
```{r}
n <- "As Michael Scott said,\n\"I'm not superstitious, but I am a little stitious.\""
print(n)
cat(n)
```
tab with \t
```{r}
t <- "this has\ta tab"
cat(t)
```
special characters and unicode with \u
```{r}
mu <- "\u00b5"
cat(mu)
```
# stringr
r package to deal with strings (grep = base R version)

```{r}
library(tidyverse)

beyonce <- "I am Beyonce, always."
str_length(beyonce)

str_length("I am Beyonce,\n\t\u2022 always.")

```
```{r}
x <- "Would I rather be feared or loved?"
y <- "Easy. Both."
z <- "I want people to be afraid of how much they love me."

str_c(x, y, z, sep = " ")
str_c("Where", "are", "the", "turtles?!", sep = " ")

```
str_sub()
```{r}
bankrupt <- "I... declare... Bankruptcy!"
str_sub(bankrupt, start = 4, end = 10)
str_sub(bankrupt, start = 6, end = 12)
str_sub(bankrupt, start = -10)


phone <- "800-800-8553"
str_sub(phone, start = 1, end = 3)
str_sub(phone, start = -4)

str_sub(bankrupt, end = 1) <- "We"


```
str_replace()
```{r}
bankrupt
str_replace(bankrupt, pattern = "We", replacement = "I")
```



----------------------------

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
