---
title: "Chpt. 14: Strings & Regex (regular expressions)"
output: html_notebook
html: TRUE
---

```{r}
library(tidyverse)
library(htmlwidgets)
```


```{r}
x <- "this is a string"
x
class(x)
y <- "3"
class(y)
as.numeric(y)
as.numeric(x) #doesn't work! 

z <- c("I", "am", "a", "string", "vector")
z
z[1:2]
z[2:5]
z[-5]

#ugly
print(z)
#simplified and interpreted (pretty)
cat(z)

```

escape special characters with \
```{r}
a <- "As Michael Scott said, \"I'm not superstitious, but I am a little stitious.\""
print(a)
cat(a)
```
line break with \n
```{r}
n <- "As Michael Scott said,\n\"I'm not superstitious, but I am a little stitious.\""
print(n)
cat(n)
```
tab with \t
```{r}
t <- "this has\ta tab"
cat(t)
```
special characters and unicode with \u
```{r}
mu <- "\u00b5"
cat(mu)
```
# stringr
r package to deal with strings (grep = base R version)

```{r}
library(tidyverse)

beyonce <- "I am Beyonce, always."
str_length(beyonce)

str_length("I am Beyonce,\n\t\u2022 always.")

```
```{r}
x <- "Would I rather be feared or loved?"
y <- "Easy. Both."
z <- "I want people to be afraid of how much they love me."

str_c(x, y, z, sep = " ")
str_c("Where", "are", "the", "turtles?!", sep = " ")

```
str_sub()
```{r}
bankrupt <- "I... declare... Bankruptcy!"
str_sub(bankrupt, start = 4, end = 10)
str_sub(bankrupt, start = 6, end = 12)
str_sub(bankrupt, start = -10)


phone <- "800-800-8553"
str_sub(phone, start = 1, end = 3)
str_sub(phone, start = -4)

str_sub(bankrupt, end = 1) <- "We"


```
str_replace()
```{r}
bankrupt
str_replace(bankrupt, pattern = "We", replacement = "I")

#replace the first match
str_replace(phone, pattern = "800-", replacement = "")
#replace all matches
str_replace_all(phone, pattern = "800-", replacement = "")
```
Regular Expressions

```{r}
fruit <- c("Apple", "strawberry", "Banana", "Pear", "Blackberry", "*berry")
str_view(fruit, "an")
str_view_all(fruit, "an")
str_view(fruit, "berry")

#. <- any character
str_view(fruit, ".berry")

# \w or [:alpha:] <- any alpha chsaracter
str_view(fruit, "[:alpha:]berry")
str_view(fruit, "\\wberry")

# \\* <- a literal asterisk
str_view(fruit, "\\*berry")


```


Anchoring
beginning: ^
end: $

```{r}
str_view(fruit, "^B") #start with B
str_view(fruit, "a$") #ends with a

#replace all 4 letter words that begin with a with "foo
x <- c("apple", "barn", "ape", "cart", "alas", "pain", "ally")
str_replace(x, pattern = "^a...$", replacement = "foo")
str_view(x, pattern = "^a...$")


```
```{r}
phones <- c("Abba: 555-1234", "Anna: 555-0987", "Andy: 555-7654")

# \d <- digit
str_view(phones, pattern = "\\d\\d\\d-\\d\\d\\d\\d")

# \s [:space:] <- whitespace
str_view(phones, pattern = "\\s")

```
Alternates
```{r}
str_view(phones, "A..a")
str_view(phones, "A[bn][bn]a")
str_view(phones, "A[^b]") # ^ inside [] means not
str_view(phones, "An(na|dy)")

str_view(phones, "(?i)an(na|dy)") #(?i <- ignore case)

```


Quantifiers

?, 0 or 1 time
+, 1 or more times
*, 0 or more times

```{r}
str_view(phones, "\\d+")
str_view(phones, "\\d?")

x <- c("A", "AA", "AAA", "AAAA", "B", "BB")
str_view(x, "^A?")
str_view(x, "^A+")
str_view("color and colour", "colou?r")

```
{n}, exactly n
{n,}, n or more
{0,n}, at most, m
{n,m}, between n and m times

```{r}
str_view(phones, "\\d{3}-\\d{4}")
str_view(phones, "\\d*-\\d*")

```
Regular expressions are greedy
```{r}
love <- "Would I rather be feared or loved? Easy. Both. I want people to be afraid of how much they love me."

str_view(love, "love.*")
str_view(love, "love.?")

str_replace(love, pattern = "love[:alpha:]?", replacement = "x")
str_replace_all(love, pattern = "love[:alpha:]?", replacement = "x")

#replace love or loved with x
```

```{r}
#replace strings with 3 consonants in a row at the start of a word
x1 <- c("string", "priority", "value", "distinction")
str_view(x1, "^[^aeiou]{3}")
```
```{r}
cause <- "I have a cause. It is beCAUSE I hate him."
str_to_lower(cause)
str_to_upper(cause)
str_to_sentence(cause)
str_to_title(cause)
```
```{r}
library(Lahman)
data("People")
People <- People |> 
  as_tibble()

People |> 
  filter(str_detect(nameFirst, "^Jo(e|hn)$")) |> 
  select(nameFirst, nameLast) |> 
  head()
```


----------------------------

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
