---
title: "Building the Past"
author: "Austin Mangelson"
format: 
  html:
    code-fold: true
    code-tools: true
    embed-resources: true
editor: visual
execute: 
  warning: FALSE
  message: FALSE
---

Between 2007 and 2010, the United States faced one of the [most severe housing market collapses](https://en.wikipedia.org/wiki/Subprime_mortgage_crisis) in its history. The ripple effects of the mortgage crisis led to a dramatic downturn in residential construction activities, profoundly impacting single-family housing permits across the nation.

In this report, we'll be focusing on single-family housing permits in1980 and in 2010.

A nationwide decrease in single-family home permit values can be a signal for many things, including economic slowdown or recession, affordability issues, and a shifting market.

## Across the U.S.

```{r setup, echo=FALSE, results='hide'}

devtools::install_github("hathawayj/buildings")
library(rnaturalearth)
library(tidyverse)
library(devtools)
library(usethis)
library(buildings)
library(USAboundaries)
library(stringr)
library(sf)
library(DT)
library(ggrepel)

```

```{r data input and tidying, echo=FALSE, results='hide'}

world_states <- ne_states()

states <- maps::map("state", plot = FALSE, fill = TRUE) |>
  st_as_sf() |> 
  mutate(ID = str_to_title(ID)) |> 
  rename(name = "ID")

us_states <- world_states |> 
  filter(geonunit == "United States of America", 
         !woe_name %in% c("Alaska", "Hawaii"))

permits_us <- buildings::permits |> 
  rename(FIPS.state = state,
         FIPS.county = county)

permits_us_single_fam <- buildings::permits |> 
  rename(FIPS.state = state,
         FIPS.county = county) |> 
  filter(variable == "Single Family")

permits_id <- buildings::permits |> 
  filter(StateAbbr == "ID") |> 
  rename(FIPS.state = state,
         FIPS.county = county)

permits_id_single_fam <- buildings::permits |> 
  filter(StateAbbr == "ID",
         variable == "Single Family") |> 
  rename(FIPS.state = state,
         FIPS.county = county)

us_county_data <- st_read("../data/spatial_idaho/shp/County-AK-HI-Moved-USA-Map.shp") |> 
  select(StateFIPSN, CntyFIPSN, StateName, CntyName, geometry) |> 
  rename(FIPS.county = "CntyFIPSN",
         FIPS.state = "StateFIPSN")

id_county_data <- us_county_data |> 
  filter(StateName == "Idaho") |> 
  st_transform(us_county_data, crs = "NAD27")

merged_id_data <- permits_id_single_fam |> 
  left_join(id_county_data, by = c("FIPS.county", "FIPS.state"))

merged_us_data <- permits_us_single_fam |> 
  left_join(us_county_data, by = c("FIPS.county", "FIPS.state"))

```

::: panel-tabset
## Plot

```{r permit decline plot}

us_permit_values_adj <- merged_us_data |> 
  mutate(StateName = str_to_title(StateName)) |> 
  rename(name = "StateName") |> 
  select(!StateAbbr, !CntyName) |> 
  group_by(name, year, FIPS.state) |> 
  summarise(total_value = sum(value, na.rm = TRUE)) |> 
  filter(year %in% c("1980", "2010")) |> 
  pivot_wider(names_from = year, values_from = total_value) |> 
  mutate(`1980_adj` = (`1980` * 2.785),
         amnt_change = ((`2010` - `1980_adj`) / `1980_adj`)) |> 
  left_join(states, by = "name") |> 
  na.omit() |> 
  rename(geometry = "geom")

ggplot() +
  geom_sf(data = us_permit_values_adj, aes(geometry = geometry, fill = amnt_change)) +
  scale_fill_gradient(low = "#FF2500", high = "#FFFC74", labels = scales::percent_format()) + 
  theme_void() +
  labs(
    title = "Single Family Housing Permit Value Decline in US \nfrom 1980 to 2010",
    subtitle = "Adjusted for inflation according to BLS",
    fill = "Percent Change"
  ) +
  coord_sf(crs = "NAD27")

```

## Data

```{r data}

datatable(us_permit_values_adj)

```
:::

As shown above, the value of single-family housing permits declined significantly in 2010 as compared to 1980, with states like California, Florida, Oregon, Rhode Island, and Connecticut seeing over an 80% decrease in permit values.

## Looking at Idaho

Idaho, like many other states, experienced a noticeable decrease in new home construction permit values. This drop reflects the broader economic troubles of the time, where obtaining loans became harder, and fewer new homes were being built.

::: panel-tabset
## Plot

```{r ID permit decline plot}

id_permit_values_adj <- merged_id_data |> 
  mutate(StateName = str_to_title(StateName)) |> 
  rename(name = "StateName") |> 
  select(!StateAbbr, !CntyName) |> 
  group_by(countyname, year, FIPS.state) |> 
  summarise(total_value = sum(value, na.rm = TRUE)) |> 
  filter(year %in% c("1980", "2010")) |> 
  pivot_wider(names_from = year, values_from = total_value) |> 
  mutate(`1980_adj` = (`1980` * 2.785),
         percent_change = ((`2010` - `1980_adj`) / `1980_adj`)) |> 
  rename(CntyName = "countyname") |> 
  left_join(id_county_data, by = "CntyName") |> 
  select(-FIPS.state.x, -FIPS.state.y, -StateName) |> 
  mutate(percent_change = if_else(percent_change > 0, NA_real_, percent_change)) |> 
  mutate(CntyName = str_sub(CntyName, 1, -8))

ggplot() +
  geom_sf(data = id_permit_values_adj, aes(geometry = geometry, fill = percent_change)) +
  scale_fill_gradient(low = "#FF2500", high = "#FFFC74", na.value = "grey85", labels = scales::percent_format()) + 
  theme_void() +
  labs(
    title = "Single Family Housing Permit Value Decline in Idaho \nfrom 1980 to 2010",
    subtitle = "Adjusted for inflation according to BLS",
    fill = "Percent Change"
  ) +
  coord_sf(crs = "NAD27")

```

## Labeled

```{r labeled ID permit decline plot}

ggplot() +
  geom_sf(data = id_permit_values_adj, aes(geometry = geometry, fill = percent_change)) +
  scale_fill_gradient(low = "#FF2500", high = "#FFFC74", na.value = "grey85", labels = scales::percent_format()) + 
  theme_void() +
  geom_label_repel(data = id_permit_values_adj, aes(geometry = geometry, label = CntyName), 
                   size = 2, nudge_x = -0.01, nudge_y = 0.2, 
                   stat = "sf_coordinates") +
  labs(
    title = "Single Family Housing Permit Value Decline in Idaho \nfrom 1980 to 2010",
    subtitle = "Adjusted for inflation according to BLS",
    fill = "Percent Change"
  ) +
  coord_sf(crs = "NAD27")

```

## Data

```{r ID data}

datatable(id_permit_values_adj)

```
:::

## Additional Critical Factors

Beyond permit values, other critical factors and indicators to consider when analyzing the mortgage crisis include (but aren't limited to):

-   foreclosure rates

-   loan prices over time

-   housing supply

-   credit conditions
