---
title: "Chpt. 1 - Data Visualization"
author: "Austin Mangelson"
format: html
editor: visual
---

```{r}
library(tidyverse)
library(palmerpenguins)
library(ggthemes)
```

**Tidy data?** Each value has it's own cell, each variable it's own column, and each observation it's own row

## First Steps

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
  ) +
  geom_point(mapping = aes(color = species, shape = species)) +
  geom_smooth(method = "lm") +
  labs(
    title = "Body mass and flipper length",
    subtitle = "Dimensions for Adelie, Chinstrap, and Gentoo Penguins",
    x = "Flipper length (mm)", y = "Body mass (g)",
    color = "Species", shape = "Species"
  ) +
  scale_color_colorblind()
```

### Exercises

1.  How many rows are in `penguins`? How many columns?

rows = 344, columns = 8

2.  What does the `bill_depth_mm` variable in the `penguins` data frame describe? Read the help for [`?penguins`](https://allisonhorst.github.io/palmerpenguins/reference/penguins.html) to find out.

bill_depth_mm = "a number denoting bill length (millimeters)

3.  Make a scatterplot of `bill_depth_mm` vs. `bill_length_mm`. That is, make a scatterplot with `bill_depth_mm` on the y-axis and `bill_length_mm` on the x-axis. Describe the relationship between these two variables.

There seems to be a positive, moderate correlation between bill depth and length among each species. That is penguins with a longer bill length are more likely to have a deeper bill depth. However, this relationship isn't as strong as flipper length and body mass.

```{r}
ggplot(
  data = penguins, 
  mapping = aes(x = bill_length_mm, y = bill_depth_mm)
) +
  geom_point(aes(color = species))
```

4.  What happens if you make a scatterplot of `species` vs. `bill_depth_mm`? What might be a better choice of geom?

This plot is unreadable because it plots categorical data as continuous data.

```{r}
ggplot(penguins, aes(x = species, y = bill_depth_mm)) +
  geom_point()
```

A better geom choice would be geom_boxplot().

```{r}
ggplot(penguins, aes(x = species, y = bill_depth_mm)) +
  geom_boxplot()
```

5.  Why does the following give an error and how would you fix it?

```{r}
ggplot(data = penguins) + 
  geom_point()
```

This gives an error because there is no x or y argument in the aesthetic. To fix this, choose 2 continuous variable to compare. The following code fixes the problem:

```{r}
ggplot(data = penguins, aes(x = bill_length_mm, y = body_mass_g)) + 
  geom_point()
```

6.  What does the `na.rm` argument do in [`geom_point()`](https://ggplot2.tidyverse.org/reference/geom_point.html)? What is the default value of the argument? Create a scatterplot where you successfully use this argument set to `TRUE`.

|                                                                                                                     |
|---------------------------------------------------------------------------------------------------------------------|
| If `FALSE`, the default, missing values are removed with a warning. If `TRUE`, missing values are silently removed. |

The na.rm argument removes or includes a warning for any NA values not included in the plot.

```{r}

ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(na.rm = TRUE)
```

7.  Add the following caption to the plot you made in the previous exercise: “Data comes from the palmerpenguins package.” Hint: Take a look at the documentation for [`labs()`](https://ggplot2.tidyverse.org/reference/labs.html).

```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(na.rm = TRUE) +
  labs(
    caption = "Data comes from the palmerpenguins package."
  )
```

8.  Recreate the following visualization. What aesthetic should `bill_depth_mm` be mapped to? And should it be mapped at the global level or at the geom level?

bill_depth_mm should be mapped to the color aesthetic and should be mapped at the local level (geom_point()).

```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = bill_depth_mm)) +
  geom_smooth()
```

9.  Run this code in your head and predict what the output will look like. Then, run the code in R and check your predictions.

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g, color = island)
) +
  geom_point() +
  geom_smooth(se = FALSE)
```

10. Will these two graphs look different? Why/why not?

They will look the same because the aesthetics are the same. In one they are applied at the global level, the other at the local level.

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point() +
  geom_smooth()
```

```{r}
ggplot() +
  geom_point(
    data = penguins,
    mapping = aes(x = flipper_length_mm, y = body_mass_g)
  ) +
  geom_smooth(
    data = penguins,
    mapping = aes(x = flipper_length_mm, y = body_mass_g)
  )
```

------------------------------------------------------------------------

## Visualizing Distributions

**Categorical**

1.  bar charts

```{r}
ggplot(penguins, aes(x = species)) +
  geom_bar()
```

```{r}
ggplot(penguins, aes(x = fct_infreq(species))) +
  geom_bar()
```

**Numerical** (*can be continuous or discrete*):

1.  histogram

```{r}
ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(binwidth = 200)
```

Too small...

```{r}
ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(binwidth = 20)
```

Too big...

```{r}
ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(binwidth = 2000)
```

2.  Density plot

```{r}
ggplot(penguins, aes(x = body_mass_g)) +
  geom_density()
```

### Exercises:

1.  Make a bar plot of `species` of `penguins`, where you assign `species` to the `y` aesthetic. How is this plot different?

This changes the plot from vertical bars to horizontal bars.

```{r}
ggplot(penguins, aes(y = species)) +
  geom_bar()
```

This does the same as coord_flip()

```{r}
ggplot(penguins, aes(x = species)) +
  geom_bar() +
  coord_flip()
```

2.  How are the following two plots different? Which aesthetic, `color` or `fill`, is more useful for changing the color of bars?

Fill = inside color of bars, while color = outline

```{r}
ggplot(penguins, aes(x = species)) +
  geom_bar(color = "red")
```

```{r}
ggplot(penguins, aes(x = species)) +
  geom_bar(fill = "red") 
```

3.  What does the `bins` argument in [`geom_histogram()`](https://ggplot2.tidyverse.org/reference/geom_histogram.html) do?

Bins = "Number of bins. Overridden by `binwidth`. Defaults to 30."

```{r}
ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(bins = 15)
```

4.  Make a histogram of the `carat` variable in the `diamonds` dataset that is available when you load the tidyverse package. Experiment with different binwidths. What binwidth reveals the most interesting patterns?

```{r}
ggplot(diamonds, aes(x = carat)) +
  geom_histogram(binwidth = 0.2)
```

## Visualizing Relationships

**Numerical vs. categorical**

1.  Boxplot

```{r}
ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_boxplot()
```

2.  Density plot

```{r}
ggplot(penguins, aes(x = body_mass_g, color = species)) +
  geom_density(linewidth = 0.75)
```

```{r}
ggplot(penguins, aes(x = body_mass_g, color = species, fill = species)) +
  geom_density(alpha = 0.5)
```

**Categorical vs categorical**

1.  stacked bar plots

```{r}
ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar()
```

2.  relative frequency plot

```{r}
ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(position = "fill")
```

**Numeric vs Numeric**

1.  scatterplot (most common)

```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point()
```

**3+ variables**

Incorporate colors/shapes

```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species, shape = island))
```

*Don't include too many aesthetics per plot!* Instead, facet wrap

```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species, shape = species)) +
  facet_wrap(~island)
```

### Exercises

1.  The `mpg` data frame that is bundled with the ggplot2 package contains 234 observations collected by the US Environmental Protection Agency on 38 car models. Which variables in `mpg` are categorical? Which variables are numerical? (Hint: Type [`?mpg`](https://ggplot2.tidyverse.org/reference/mpg.html) to read the documentation for the dataset.) How can you see this information when you run `mpg`?

**Categorical**: manufacturer, model, trans, dvr, fl, class

**Numerical**: displ, year, cyl, cty, hwy

2.  Make a scatterplot of `hwy` vs. `displ` using the `mpg` data frame. Next, map a third, numerical variable to `color`, then `size`, then both `color` and `size`, then `shape`. How do these aesthetics behave differently for categorical vs. numerical variables?

```{r}
ggplot(mpg, aes(x = hwy, y = displ)) +
  geom_point(aes(size = cyl))
```

3.  In the scatterplot of `hwy` vs. `displ`, what happens if you map a third variable to `linewidth`?

Nothing, because there are no lines in a geom_point()

```{r}
ggplot(mpg, aes(x = hwy, y = displ)) +
  geom_point(aes(linewidth = cyl))
```

4.  What happens if you map the same variable to multiple aesthetics?

    ```{r}
    ggplot(mpg, aes(x = hwy, y = displ)) +
      geom_point(aes(size = cyl, color = cyl))
    ```

5.  Make a scatterplot of `bill_depth_mm` vs. `bill_length_mm` and color the points by `species`. What does adding coloring by species reveal about the relationship between these two variables? What about faceting by `species`?

    ```{r}
    ggplot(penguins, aes(x = bill_depth_mm, y = bill_length_mm)) +
      geom_point(aes(color = species))
    ```

    ```{r}
    ggplot(penguins, aes(x = bill_depth_mm, y = bill_length_mm)) +
      geom_point(aes(color = species)) +
      facet_wrap(~species)
    ```

6.  Why does the following yield two separate legends? How would you fix it to combine the two legends?

Because species is mapped to both color and shape. Also because species is lowercase as a variable, but uppercase in the legend title.

```{r}
ggplot(
  data = penguins,
  mapping = aes(
    x = bill_length_mm, y = bill_depth_mm, 
    color = species, shape = species
  )
) +
  geom_point() +
  labs(color = "Species")
```

Fixed plot:

```{r}
ggplot(penguins, aes(
  x = bill_length_mm, 
  y = bill_depth_mm, 
  color = species, 
  shape = species)
) +
  geom_point() +
  labs(color = "Species",
       shape = "Species")
```

7.  Create the two following stacked bar plots. Which question can you answer with the first one? Which question can you answer with the second one?

First answers population of specific species per island. Second answers where can you find each species of penguin.

```{r}
ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(position = "fill")
ggplot(penguins, aes(x = species, fill = island)) +
  geom_bar(position = "fill")
```
